ppseq: An R Package for Sequential Predictive Probability Monitoring

### 2022-11-09 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
change back to com
update package paper

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent trials of targeted treatments have included expansion cohorts with sample sizes far exceeding those in traditional early phase trials of chemotherapeutic agents. The enlarged sample sizes raise ethical concerns for patients who enroll in clinical trials, and emphasize the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials of binary endpoints using sequential futility monitoring based on Bayesian predictive probability. Trial designs can be compared using interactive plots and selected based on measures of efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.me/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zabor-hobbs-kane.bib
---

@@

After obtaining results for all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options. The \CRANpkg{ppseq} package introduces two criteria to assist users in making a selection. The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to 0 type I error probability and a power of 1. To accomplish this, the accuracy Euclidean distance (AED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $AED_{km} = w_\alpha * (\alpha_{km} - 0)^2 + w_(1-\beta) * ((1-\beta)_{km}-1)^2$, where $w_\alpha$ and $w_{(1-\beta)}$ are optional weights on the type I error and power, respectively, and $\alpha_{km}$ denotes the estimated type I error and $(1-\beta)_{km}$ denotes the estimated power. The design with the smallest value of $AED_{km}$ is selected as optimal. The second criteria, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to minimal average sample size under the null and maximal average sample size under the alternative. To accomplish this, the efficiency Euclidean distant (EED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $EED_{km} = w_{\bar{N}_{H_0}} * (\bar{N}_{H_0km} - min(\bar{N}_{H_0}))^2 + w_{\bar{N}_{H_1}} * (\bar{N}_{H_1km}-max(\bar{N}_{H_1}))^2$, where $w_{\bar{N}_{H_0}}$ and $w_{\bar{N}_{H_1}}$ are optional weights on the average sample size under the null and alternative, respectively, $\bar{N}_{H_0km}$ and $\bar{N}_{H_1km}$ denote the average sample sizes under the null and alternative, respectively, and $min(\bar{N}_{H_0})$ and $max(\bar{N}_{H_1})$ denote the minimum average sample size under the null and the maximum average sample size under the alternative alternative, respectively, across all combinations of $k$ and $m$. The design with the smallest value of $EED_{km}$ is selected as optimal. The `optimize_design()` function returns a list that contains the details of each of the two optimal designs.

@@

```{r message = FALSE, fig.height = 6, fig.width = 7, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot. The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
dtp <- plot(two_sample_decision_tbl, plotly = FALSE)

>>>>

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent trials of targeted treatments have included expansion cohorts with sample sizes far exceeding those in traditional early phase trials of chemotherapeutic agents. The enlarged sample sizes raise ethical concerns for patients who enroll in clinical trials, and emphasize the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials of binary endpoints using sequential futility monitoring based on Bayesian predictive probability. Trial designs can be compared using interactive plots and selected based on measures of efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zabor-hobbs-kane.bib
---

@@

After obtaining results for all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options. The \CRANpkg{ppseq} package introduces two criteria to assist users in making a selection. The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to 0 type I error probability and a power of 1. To accomplish this, the accuracy Euclidean distance (AED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $AED_{km} = w_\alpha * (\alpha_{km} - 0)^2 + w_{(1-\beta)} * ((1-\beta)_{km}-1)^2$, where $w_\alpha$ and $w_{(1-\beta)}$ are optional weights on the type I error and power, respectively, and $\alpha_{km}$ denotes the estimated type I error and $(1-\beta)_{km}$ denotes the estimated power. The design with the smallest value of $AED_{km}$ is selected as optimal. The second criteria, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to minimal average sample size under the null and maximal average sample size under the alternative. To accomplish this, the efficiency Euclidean distant (EED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $EED_{km} = w_{\bar{N}_{H_0}} * (\bar{N}_{H_0km} - min(\bar{N}_{H_0}))^2 + w_{\bar{N}_{H_1}} * (\bar{N}_{H_1km}-max(\bar{N}_{H_1}))^2$, where $w_{\bar{N}_{H_0}}$ and $w_{\bar{N}_{H_1}}$ are optional weights on the average sample size under the null and alternative, respectively, $\bar{N}_{H_0km}$ and $\bar{N}_{H_1km}$ denote the average sample sizes under the null and alternative, respectively, and $min(\bar{N}_{H_0})$ and $max(\bar{N}_{H_1})$ denote the minimum average sample size under the null and the maximum average sample size under the alternative alternative, respectively, across all combinations of $k$ and $m$. The design with the smallest value of $EED_{km}$ is selected as optimal. The `optimize_design()` function returns a list that contains the details of each of the two optimal designs.

@@

```{r message = FALSE, fig.height = 6, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot. The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
dtp <- plot(two_sample_decision_tbl, plotly = FALSE)
